Efficient Computation of Extended Sources

### 2021-05-04 repro/papers/sseprecond/paper.tex
interim

The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
%A typical nonlinear least squares formulation is:
%\begin{equation}
%\label{eqn:ols}
%\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2.
%\end{equation}

%As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''. Source extension is one approach to avoiding this problem. It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing the source to have more degrees of freedom than is permitted by a faithful  model model of the seismic experiment, and constraining these additional degrees of freedom by means of an additive penalty modifying the probem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } \bf{c}, \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed) characteristics of the source function - its null space consists of feasible source models.

This paper presents a numerically efficient approach to solving the {\em source subproblem}
\begin{equation}
\label{eqn:esis}
\mbox{given } \bf{c}, \mbox{ choose } \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
\end{equation}

The penalty operator $A$ is assumed linear in this work, so the source subproblem is a linear least squares problem. Under some additional assumptions to be described below, I shall show how to construct an accurate approximate solution operator for problem \ref{eqn:esis}. This approximate solution operator may be used to accelerate Krylov space methods for the solution of the source subproblem \ref{eqn:esis}. Numerical examples suggest the effectiveness of this acceleration.

\section{Surface Sources and Solutions}

For acoustic wave physics, the coefficient vector $c$ is the pair $\bf{c}=(\kappa,\rho)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field). The wave operator $L[\bf{c}]$ is given by the linear acoustics system:
\begin{equation}
\label{eqn:aweop}
L[c]u = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[c] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

Most of what follows is valid for any space dimension $n >0$. Examples
later in this paper will use $n=2$ for computationally convenience. In
the mathematical development that follows, I will set $n=3$.

As mentioned earlier, in this paper the coefficients $\bf{c}$ is regarded as defined throughout space $\bR^d$, the state vector $\bu$ throughout space-time $\bR^{d+1}$.

The idealized ``crosswell'' configuration involves sources confined to
$z=z_s$. Such sources are combinations of constitutive law defects and
loads normal to the surface, that is, right-hand sides in the system
$L[c]\bf{u}=\bf{f}$ of the form $\bf{f}(\bx,t) =
(h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
defect $h_s$ and normal force $f_s$ and $\bf{e}_z=(0,0,1)^T$. With the choice $L[\bf{c}]$ given in \ref{eqn:aweop}, the causal wave system $L[\bf{c}]\bu^+=\bff$ takes the form
\begin{eqnarray}
\label{eqn:awep}
\frac{1}{\kappa}\frac{\partial p^{+}}{\partial t} & = & - \nabla \cdot \bv^{+} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
\bv^{+} & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}

Similarly, the anti-causal wave system $L[c]u^-=f$ takes the form
\begin{eqnarray}
 \label{eqn:awem}
\frac{1}{\kappa}\frac{\partial p^{-}}{\partial t} & = & - \nabla \cdot \bv^{-} +
h_s \delta(z-z_s), \nonumber \\
\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
\bv^{-} & = & 0 \mbox{ for } t \gg 0.
\end{eqnarray}

\section{Operators and Adjoints}
All of the operators discussed in this section depend on the
coefficient array $\bf{c}$, which I will suppress from the notation.
Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
respectively. The causal and anticausal modeling operators
${\cal S}^{+}_{z_s,z_r}$ and ${\cal S}^-_{z_s,z_r}$ are defined in
terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
\ref{eqn:awep} and \ref{eqn:awem} by
\begin{eqnarray}
  {\cal S}^{+}_{z_s,z_r}(h_s,f_s) & = & (P_rp^+,P_r v_z^+)|_{z=z_r},
                                        \nonumber\\
  {\cal S}^{-}_{z_s,z_r}(h_s,f_s) & = & (P_rp^-,P_r v_z^-)|_{z=z_r}.
  \label{eqn:fwd}
\end{eqnarray}
The subscript signifies that sources are located on $z=z_s$, the
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed in the discussion below.

The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
the adjoint state method, in this case a by-product of the
conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
  p^- + f_r \bf{e_z}) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
\]
whence
\begin{equation}
\label{eqn:sadj}
 ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
\end{equation}

The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with
$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
\begin{equation}
  \label{eqn:trsadj}
  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
\begin{equation} 
  \label{eqn:trtr}
 ({\cal S}^+_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^+{\cal R}=
 -{\cal S}^-_{z_r,z_s}.
\end{equation}

>>>>

%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
A simple nonlinear least squares formulation is:
\begin{equation}
\label{eqn:ols}
\mbox{choose } \bf{c} \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2.
\end{equation}
Practical optimization formulations typically augment the objective in
\ref{eqn:ols} by additive penalties or other constraints.

As is well-known, local optimization methods are the only feasible
approach given the dimensions of a typical instance of \ref{eqn:fwi},
and those have a tendency to stall due to ``cycle-skipping''. Source
extension is one approach to avoiding this problem. It consists in
imposing the wave equation as a soft as opposed to hard constraint, by
allowing the source field $\bff$ to have more degrees of freedom than
is permitted by a faithful model model of the seismic experiment, and
constraining these additional degrees of freedom by means of an
additive penalty modifying the probem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } \bf{c}, \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
literature on source extension methods, describing a variety of
methods to add degrees of freedom to physical source model. The present paper
concerns {\em surface source extension}: physical sources are
presumed to be concentrated at points $\bx_s$ in space, whereas their extended
counterparts are permitted to spread energy over surfaces containing
the physical source locations. A simple choice for the penalty
operator $A$ is then multiplication by the distance $|\bx-\bx_s|$ to the physical
source location:
\begin{equation}
  \label{eqn:penop}
  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
\end{equation}
I shall use this choice of penalty operator whenever a specific choice
is necessary in the development of the theory below.

This paper presents a numerically efficient approach to solving the {\em source subproblem}
\begin{equation}
\label{eqn:esis}
\mbox{given } \bf{c}, \mbox{ choose } \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
\end{equation}
Solution of this subproblem is an essential component of {\em variable
  projection} algorithms for solution of the nonlinear inverse problem
\ref{eqn:esi}. Variable projection is not merely a convenient choice
of algorithm for this purpose: it is in some sense essential, see for
example \cite{Symes:SEG20}. It replaces the nonlinear
least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
be solved iteratively. Each iteration involves solution of the
subproblem \ref{eqn:esis}. Therefore efficient solution of the
subproblem is essential to efficient solution of the nonlinear problem
via variable projection.

The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
subproblem is a linear least squares problem. Under some additional
assumptions to be described below, I shall show how to construct an
accurate approximate solution operator for problem
\ref{eqn:esis}. This approximate solution operator may be used to
accelerate Krylov space methods for the solution of the source
subproblem \ref{eqn:esis}. Numerical examples suggest the
effectiveness of this acceleration.

\section{Operators and Adjoints}

For acoustic wave physics, the coefficient vector is
$\bf{c}=(\kappa,\rho)$, with components bulk modulus $\kappa$ and
density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
pressure $p$ (a scalar space-time field) and particle velocity $\bv$
(a vector space-time field). The wave operator $L[\bf{c}]$ is:
\begin{equation}
\label{eqn:aweop}
L[\bf{c}]\bf{u} = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[\bf{c}] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

Most of what follows is valid for any space dimension $n >0$. The
coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
$\bR^n$, the state vector $\bu$ throughout space-time
$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
$n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
coordinate vector, and refer to the third (vertical) coordinate of
particle velocity as $v_z$. Examples
later in this paper will use $n=2$ for computational convenience.

Since all of the operators in the discussion that follows depend on
$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.

The extended ``crosswell'' configuration involves sources confined to
a spatial hypersurface. For convenience, choose this to be the hyperplane
$\{(x,y,z,t): z=z_s\}$. Such sources are combinations of constitutive law defects and
loads normal to the surface, that is, right-hand sides in the system
$L\bu=\bff$ of the form $\bff(\bx,t) =
(h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
defect $h_s$ and normal force $f_s$  ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in \ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$ takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
systems to follow, I will use the shorthand
\[
  p^+ = 0 \mbox{ for } t \ll 0
\]
to mean
\[
  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t \ge
  T.
\]
Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$. Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
respectively. The causal/anti-causal modeling operators
${\cal S}^{\pm}_{z_s,z_r}$ are defined in
terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
\ref{eqn:awepm} by
\begin{equation}
  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)  = (P_rp^{\pm},P_r v_z^{\pm}),
  \label{eqn:fwd}
\end{equation}
The subscript signifies that sources are located on $z=z_s$, the
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed in the discussion below.

It follows from the adjoint state method (see Derivations section
below for details) that
\begin{equation}
  \label{eqn:sadj1}
  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
\end{equation}

Define $R$ to be the {\em time-reversal operator} on functions of
space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
  acoustic field time-reversal operator}
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then 
\begin{equation}
  \label{eqn:trsadj}
  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that

\begin{equation} 
  \label{eqn:trtr}
 ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
 -{\cal S}^{\mp}_{z_r,z_s}.
\end{equation}

Define $\Pi_i, i=0,1$ to be the projection on the first,
respectively second, component of a vector in $\bR^2$. Define the
forward modeling operator from pressure source to pressure trace by
\begin{equation}
  \label{eqn:sdef}
  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
\end{equation}
and the forward modeling operator from velocity source (normal force)
to velocity trace by 
\begin{equation}
  \label{eqn:vdef}
  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
\end{equation}
The relation \ref{eqn:trtr} implies that
\begin{eqnarray}
  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
                        &=& R V^{\pm}_{z_r,z_s}R.
                            \label{eqn:trtrcomp}
\end{eqnarray}

\section{Fields, Traces and Sources}
In this section, $(p^{\pm}_0,\bv^{\pm}_0)$ denotes an acoustic field,
continuous in the region $z_s \le z $, obeying the homogeneous wave
equation in the interior $z_s < z $:
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_0}{\partial t} & = & - \nabla \cdot \bv^{\pm}_0, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_0}{\partial t} & = & - \nabla
                                                    p^{\pm}_0, \nonumber \\
  p^{\pm}_0 & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
  \bv^{\pm}_0 & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}
I will show how to view $(p^{\pm}_0,\bv^{\pm}_0)$ as the $z>z_s$ part
of a solution of the system \ref{eqn:awepm} in two different ways,
corresponding to two different extensions to the whole space. An
important relation between the components of the operator
${\cal S}^{\pm}_{z_s,z_r}$ follows (equation \ref{eqn:lamnull} below).

Given a function $\phi$ on $z=z_s$, define $(p^{\pm}_1,\bv^{\pm}_1)$
as the solution in the two half-spaces $z <z_s, z>z_s$ of
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
  \lim_{z \rightarrow z_s} P_sp^{\pm}_1& =& \phi.
\end{eqnarray}
By construction, $p^{\pm}_1$ is continuous across $z=z_s$. The
discontinuity $[v^{\pm}_{1,z}]|_{z=z_s}$ is a linear function of
$\phi$. Define the {\em pressure-to-source map} $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s}
\end{equation}
Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solve the
inhomogeneous system
\begin{eqnarray}
\label{eqn:sawe1}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_1}{\partial t} & = &
                                                        (\Lambda^{\pm}_{z_s}\phi)\delta(z-z_s) - \nabla \cdot \bv^{\pm}_1, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \nabla p^{\pm}_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

This is precisely the causal option of system \ref{eqn:awepm} for
$f_s=0$ and $h_s=\Lambda^{\pm}_{z_s}\phi$. From \ref{eqn:awe1} and the definition
\ref{eqn:fwd},
\begin{equation}
  \label{eqn:lamfwd}
  \phi = S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s}\phi,
\end{equation}
that is,
\begin{equation}
  \label{eqn:lamfwd1}
  S^{\pm}_{z_s,z_s} = (\Lambda^{\pm}_{z_s})^{-1}.
\end{equation}

Similarly, suppose $\psi$ is a function on $z=z_s$, and that
$(p^{\pm}_2,\bv^{\pm}_2)$ solves
\begin{eqnarray}
\label{eqn:awe2}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = & - \nabla \cdot \bv^{\pm}_2, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & - \nabla p^{\pm}_2,\nonumber \\
  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
  \lim_{z \rightarrow z_s} P_sv^{\pm}_{2,z} &=& \psi.
\end{eqnarray}
in the two half-spaces $z<z_s, z > z_s$. In this case, $v^{\pm}_{2,z}$
is continuous at $z=z_s$ by construction. Viewed as a
distribution on all of $\bR^4$, $(p^{\pm}_2,\bv^{\pm}_2)$ solve the
system \ref{eqn:awepm} with $h_s=0$, and the discontinuity in pressure
across $z=z_s$ as the inhomogeneous term in Newton's law:
\begin{eqnarray}
\label{eqn:sawe2}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = &
                                                        - \nabla \cdot \bv^{\pm}_2, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & 
                                                    [p^{\pm}_2]|_{z=z_s}{\bf e_z}\delta(z-z_s) - \nabla p^{\pm}_2,\nonumber \\
  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}
This relation defines a map from the velocity boundary data $\psi$ to
the source in the velocity equation, having units of force/volume,
similar to $\Lambda^{\pm}_{z_s}$. As it's not needed in the sequel, I
won't name this map.

The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
field $(p^{\pm},\bv^{\pm})$, at time $t$ by
\begin{equation}
  \label{eqn:defae0}
  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
\end{equation}
Then
\begin{equation}
  \label{eqn:elim}
  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
\end{equation}
That is, the value of the quadratic form defined by
$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
gives the total energy transferred from the source to the
acoustic field over time. Since $E$ is itself a positive definite
quadratic form in the acoustic field, it follows that $\pm
\Lambda^{\pm}_{z_s}$ is positive semi-definite.

A closer relation between traces and sources follows from the
geometric optics analysis of the half-space solution, presented
below. I will show that, for propagating wave data,
\begin{equation}
  \label{eqn:odd1}
  \lim_{z\rightarrow z_s^+} v_{1,z} \approx - \lim_{z\rightarrow z_s^-}
  v_{1,z}
\end{equation}
Similarly,
\begin{equation}
  \label{eqn:odd2}
  \lim_{z\rightarrow z_s^+} p_{2} \approx - \lim_{z\rightarrow z_s^-}
  p_{2}
\end{equation}
for the velocity field solving \ref{eqn:awe1}, or equivalently
\ref{eqn:awe1}. Here ``$\approx$'' means in the sense of high
frequency asymptotics, that is, that the difference between the two
sides is relatively smooth, hence small if the data is highly
oscillatory. Similarly, for the pressure field solving \ref{eqn:awe2}
or \ref{eqn:sawe2}, the jump at $z=z_s$ is twice the limiting value
(trace) from above, up to a smooth error. Therefore
\begin{eqnarray}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s} &\approx& 2
                              \lim_{z\rightarrow z_s}v^{\pm}_{1,z},\\  \label{eqn:tracejump1}
  [p^{\pm}_2]|_{z=z_s} &\approx& 2 \lim_{z\rightarrow z_s^+}p^{\pm}_{2},\\  \label{eqn:tracejump2}
\end{eqnarray}

These apparently harmless relations have a profound consequence,
because of a final step: identify the boundary data $\phi$ and $\psi$,
so far arbitrary, as the traces of $p^{\pm}_0$ and $v^{\pm}_{0,z}$ on $z=z_s$,
that is, set
\begin{eqnarray}
  \phi & = & P_s p^{\pm}_0, \nonumber \\
  \psi & = & P_s v^{\pm}_{0,z}.
             \label{eqn:bv0}
\end{eqnarray}
Since the traces on $z=z_s$ of the pressure fields $p^{\pm}_0$ and $p^{\pm}_1$ are
the same, and both are pressure components of solutions of the same homogeneous
acoustic system in $z>z_s$, the standard energy identity implies that
$(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_1,\bv^{\pm}_1)$ in
$z > z_s$. Similarly, $P_sv^{\pm}_{0,z}$ and $P_sv^{\pm}_{2,z}$ are
the same, whence $(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_2,\bv^{\pm}_2)$ in
$z>z_s$.

Therefore the assumption \ref{eqn:bv0} implies that
\begin{equation}
  \label{eqn:jump0}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]_{z=z_s} \approx 2P_s v^{\pm}_{0,z}  = 2\psi,
\end{equation}
and
\begin{equation}
  \label{eqn:jump1}
[p_2^{\pm}]_{z=z_s} \approx  2P_s p^{\pm}_{0}= 2 \phi.           
\end{equation}
In the notation of the last section, set $2 \phi = 2P_sp^{\pm}_0= f_s,
2\psi = 2P_s v^{\pm}_{0,z}= h_s$. Then relation \ref{eqn:jump0}
implies that
\begin{equation}
  \label{eqn:lamfh}
  \Lambda^{\pm}_{z_s}f_s \approx h_s.
\end{equation}

From the definitions in the last section and the system \ref{eqn:sawe1},
\begin{equation}
  \label{eqn:pident0}
P_r p^{\pm}_0 = \Pi_0{\cal S}^{\pm}(h_s,0)^T = \Pi_0{\cal
  S}^{\pm}\Pi_0^Th_s,
\end{equation}
and
\begin{equation}
  \label{eqn:pident1}
P_r p^{\pm}_0 \approx \Pi_0{\cal S}^{\pm}_{z_s,z_r} (0,f_s)^T = \Pi_0{\cal
  S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s,
\end{equation}
Similarly, from the relations above and \ref{eqn:sawe2},
\begin{equation}
  \label{eqn:vident0}
  P_r  v^{\pm}_{0,z}  \approx \Pi_1{\cal S}^{\pm}_{z_s,z_r}(h_s,0)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_0^Th_s 
\end{equation}
and
\begin{equation}
  \label{eqn:vident1}
  P_r  v^{\pm}_{0,z}  = \Pi_1{\cal S}^{\pm}_{z_s,z_r}(0,f_s)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s 
\end{equation}
Together with \ref{eqn:lamfh}, these equations imply that
\begin{equation}
  \label{eqn:lamnull}
{\cal  S}_{z_s,z_r}^{\pm}P_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}P_1^T.
\end{equation}
Consequently, ${\cal S}_{z_s,z_r}$ has an ``approximate null space'',
in the sense that its composition with the block operator
$(\Lambda^{\pm}_{z_s},-I)^T$ is asymptotically negligible.

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. An accurate computation is available via the
computation of a {\em parametrix} (high frequency approximate inverse)
for $S^{\pm}_{z_s,z_r}$. This parametrix also justifies the use of the
symmetric part of $\Lambda^{\pm}_{z_s}$ as a preconditioner. In this
section I will explain the parametrix construction, and also show how
to extract the symmetric part of $\Lambda^{\pm}_{z_s}$.

The key ingredient in this {\em time reversal} construction is a
{\em local energy decay} assumption. Define the local energy in the
acoustic field $(p,\bv)$ at
time $t$ in a region $\Omega \subset \bR^3$ by imiting the integration
in the definition of total energy \ref{eqn:defae0} to $\Omega$:
\begin{equation}
  \label{eqn:defaeloc}
  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
\end{equation}
